Event History Analysis for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows

### 2025-06-20 manuscript/reviews/ResponseReview1.Rmd

```{r packages and data, include=F }

```

# Editor

> 15-Mar-2025
>
> Dear Dr. Panis:
>
> Thank you for submitting your Tutorial  entitled "Event History Analysis for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows" . I have now received input from three experts in the field, and you will find the reviews below this message and attached. As you will see, all three reviewers believe your paper can make a strong contribution to the literature. Prior to considering the reviewers' input, I read your paper to form an independent opinion. I share the positive sentiments about this work and invite you to revise and resubmit your manuscript for further consideration.

@@

> The reviews are through and thoughtful, and I will not reiterate all the details here. For my part, I would like your revision to focus on accessibility. Your paper is a bit challenging-- it's long and a little tough to get through. What can make it easier for less familiar readers? Perhaps more of a basic set up earlier? A glossary defining specific terms (e.g., what is an event? how is it defines? censoring? hazards vs. survival curves, conditional probabilities, etc.)? Also, clarify when readers need to shift from frequentist to Bayesian model... and why? The main consideration here is that wish to make the paper much more accessible-- from start to finish-- for readers who are novices and entirely unfamiliar with this area. This can be done any number of ways, and R3 also includes some suggestions for making these reviews. 
I encourage you to have the revised paper read and pre-reviewed by colleagues who are novices in this area with the explicit goal of making the paper accessible to them-- this will help you hit the more for our broad readership.

> If you choose to submit a revision, please include include a letter detailing your point-by-point responses to each reviewer comment and indicating how you changed the manuscript to address them. The revised manuscript may undergo further peer review. We ask that you submit your revision within three months. Please let us know if you will not be able to meet this deadline.

> Sincerely,

We tried to focus on accessibility in the revision of our Tutorial.

### 2025-06-20 manuscript/manuscript.Rmd
Feedback and edits on the updated manuscript.

# R references
r_refs(file = "r-references.bib", append=F) # append=F prevents multiple copies of the same citation
my_r_citation <- cite_r(file = "r-references.bib", footnote = TRUE) 
```

@@

As a simple illustration, Figure 1 shows how comparing means between two conditions conceal the shapes of the underlying RT and accuracy distributions. We simulated a RT + accuracy data set for a single subject who performed 200 trials (i.e., repeated measurements) in each of two conditions. For example, while this subject is 71 ms faster on average in condition 1 (481 ms) compared to condition 2 (552 ms), the corresponding hazard functions of response occurrence show that the effect starts in time period (400,500] or bin t=5, and is present in three consecutive time bins (i.e., for 300 ms). Similarly, while this subject makes less errors in condition 1 (86% accuracy) compared to condition 2 (64% accuracy), the conditional accuracy functions show that (a) the effect is present only for responses emitted before 400 ms, (b) erroneous responses in condition 1 are confined to a single time bin, and (c) the observed conditional accuracies (0, 1, 0.51, 0.48) are never even close to the mean accuracies.

Why does this matter for research in psychology? Compared to the aggregation of data across trials, a distributional approach offers the possibility to reveal the time course of psychological states. For example, Figure 1B shows a first state (up to 400 ms after target onset) for which the early upswing in hazard is equal for both conditions, and the emitted responses are always correct in condition 1 and always incorrect in condition 2. In a second state (400 to 500 ms), hazard is higer in condition 1, and conditional accuracies are close to .5 in both conditions. In a third state (>500 ms), the effect disappears in hazard, and all conditional accuracies are equal to 1.

For many psychological questions, such "temporal states" information can be theoretically meaningful by leading to more fine-grained understanding of psychological processes, by adding a relatively under-used dimension -- the passage of time -- to the theory building toolkit. Thus, a distributional approach permits different kinds of questions to be asked, different inferences to be made, and it holds the potential to better discriminate between different theoretical accounts of psychological and/or brain-based processes.

<!-- For instance, in examples 1-3, mean RT is always comparable between two conditions, while the distributions differ (Figure 1, left). In contrast, in examples 4-6, mean RT is lower in condition 2 compared to condition 1, but the RT distributions differ in each case (Figure 1, right). Therefore, a comparison of means would lead to a similar conclusion in examples 1-3, as well as examples 4-6, whereas a comparison of the distributions would lead to a different conclusion in every case.-->

(ref:descr-fig1-caption) Mean performance versus distributional analyses. (A) The mean RT (top) and overall accuracy (bottom) for two conditions are plotted. (B) The discrete-time hazard functions (top) and conditional accuracy functions (bottom) are plotted for the same data. The first second after target stimulus onset (time zero) is divided in ten bins of 100 ms. The first bin is (0,100], the last bin is (900,1000]. Note that the hazard and conditional accuracy estimates are plotted at the endpoint of each time bin. The definitions of discrete-time hazard and conditional accuracy are further explained in section 2. Error bars represent +/- 1 standard error of the mean (A) or proportion (B). Note that the distributional shapes are inspired by published results from interference paradigms such as priming and cueing tasks (refs). For example, if the target stimulus at time zero is preceded by a prime stimulus that can be congruent (condition 1) or incongruent (condition 2) to the target, then the distributions show that (a) the fastest responses (< 400 ms) are triggered exclusively by the prime, (b) between 400 and 500 ms, the prime-triggered response tendency is actively inhibited (leading to conditional accuracies close to .5), and (c) the slowest responses (>500 ms) are controlled by the target stimulus.

<!-- For our purposes here, it is enough to know that the distributions plotted represent the probability of an event occurring in that time bin, given that it has not yet occurred. Insets show mean response time per condition.-->

@@

<!-- For example, the distributions in Figure 1 show that the effect starts between 400 and 500 ms (in time bin 5) and is gone when the waiting time reaches 700 ms or more.

What kind of theory or theories could account for such effects? Are there new auxiliary assumptions that theories need to adopt? And are there new experiments that need to be performed to test the novel predictions that follow from these analyses? -->

## 1.2 Aims and structure of the paper

In this paper, we focus on a distributional method for time-to-event data known as discrete-time Event History Analysis (EHA), a.k.a. survival analysis, hazard analysis, duration analysis, failure-time analysis, and transition analysis [@singerAppliedLongitudinalData2003]. Our ultimate goal is twofold: first, we want to convince readers of the many benefits of using EHA when dealing with psychological RT data, and second, we want to provide a set of practical tutorials, which provide step-by-step instructions on how you actually perform a discrete-time EHA on RT data, as well as a complementary discrete-time speed-accuracy tradeoff (SAT) analysis on timed accuracy data in case of choice RT data.

<!-- We hope to show the added value of EHA for knowledge and theory building in cognitive psychology and related areas of research, such as cognitive neuroscience. Most importantly, we provide tutorials that provide step-by-step code and instructions in the hope that we can enable others to use EHA in a more routine, efficient and effective manner.-->

Even though EHA is a widely used statistical tool and there already exist many excellent reviews [e.g., @blossfeldTechniquesEventHistory2002; @hosmerAppliedSurvivalAnalysis2011; @teachmanAnalyzingSocialProcesses1983; @box-steffensmeierEventHistoryModeling2004] and tutorials [e.g., @allisonSurvivalAnalysisUsing2010; @landesIntroductionEventHistory2020], we are not aware of any tutorials that are aimed specifically at psychological RT (+ accuracy) data, and which provide worked examples of the key data processing and Bayesian multilevel regression modelling steps. From a historical perspective, it is worth noting that the development of analytical tools that can estimate or predict whether and when events will occur is not a new endeavour.
Indeed, hundreds of years ago, analytical methods were developed to predict the duration of time until people died [e.g., @Halley1693; @williammatthewmakehamLawMortalityConstruction1860].
The same logic has been applied to psychological time-to-event data, as previously demonstrated [@panisAnalyzingResponseTimes2020; XXXXX].

We first provide a brief overview of EHA to orient the reader to the basic concepts that we will use throughout the paper. However, this will remain relatively short, as this has been covered in detail before [@singerAppliedLongitudinalData2003; @allisonDiscreteTimeMethodsAnalysis1982; @allisonSurvivalAnalysisUsing2010]. Indeed, our primary aim here is to introduce the set of tutorials, which explain **how** to do such analyses, rather than repeat in any detail **why** you may do them. In section A of the Supplemental Material we explain how the data from typical RT tasks (detection, discrimination or categorization, bistable perception) are treated in EHA.

We then provide seven different tutorials, which are written in the R programming language and publicly available on our Github page (https://github.com/sven-panis/    Tutorial_Event_History_Analysis), along with all of the other code and material associated with the project. The tutorials provide hands-on, concrete examples of key parts of the analytical process, so that others can apply EHA to their own time-to-event data. Each tutorial is provided as an RMarkdown file, so that others can download and adapt the code to fit their own purposes. Additionally, each tutorial is made available as a .html file, so that it can be viewed by any web browser, and thus available to those that do not use R. Finally, the manuscript itself is written in R using the papaja package [@pap2024], which makes it computationally reproducible, in terms of the underlying data and figures.

<!--In Tutorial 1a, we illustrate how to process or "wrangle" a previously published RT + accuracy data set to calculate descriptive statistics when there is one independent variable. The descriptive statistics are plotted, and we comment on their interpretation. In Tutorial 1b we provide a generalisation of this approach to illustrate how one can calculate the descriptive statistics when using a more complex design, such as when there are two independent variables.

In Tutorial 2a, we illustrate how one can fit Bayesian multilevel regression models to RT data using the R package brms. We perform prior predictive checks, compare models, and interpret the plots of the predicted hazard functions for the selected model, and the posterior distributions of our contrasts of interest. In Tutorial 2b we fit Bayesian multilevel regression models to *timed* accuracy data to perform a micro-level speed-accuracy tradeoff (SAT) analysis, which complements the EHA of RT data for choice RT data.

In Tutorial 3a, we shortly illustrate how to fit similar multilevel regression models for RT data in a frequentist framework using the R package lme4. We then briefly compare and contrast these inferential frameworks when applied to EHA. In Tutorial 3b, we illustrate how to perform the SAT analysis in a frequentist framework.

In tutorial 4, we illustrate one approach to planning how much data to collect in an experiment using EHA. We use data simulation techniques to vary sample size and trial count per condition until a certain degree of statistical power or precision is reached.-->.

# 2. A brief introduction to event history analysis

To apply EHA, one must be able to:

1. define an event of interest that represents a qualitative change that can be situated in time (e.g., a button press, a saccade onset, a fixation offset, etc.);

@@

## 2.1 Single, repeatable, and recurrent events

While people can die only once, in experimental RT tasks the events of interest are typically repeatable. For example, in the target-present condition of a one-button detection task the participant is presented in each trial with a faint target stimulus that (s)he has to detect by pressing a button within a certain time window (e.g., 1 second from target onset). The single time-to-detection is thus repeatedly measured across multiple trials. In the parlance of EHA, in each trial the participant is "at risk" for response occurrence as long as the response has not occurred yet.

In a two-button discrimination task, the participant is presented in each trial with a target stimulus that (s)he has to categorize by pressing one of two buttons within a certain time window. This is known as a "competing risks" situation, because in each trial, the participant can transition from an idle state to either a correct response state or an incorrect response state.

In a bistable perception task, the participant is looking at an ambiguous stimulus (e.g., the duck-rabbit illusion, the Necker cube) for two minutes, for example, and asked to press a button each time when her/his perception switches from one possibe interpretation to the other possible interpretation. In this task, there are two events (percept A switches to percept B, percept B switches to percept A) that can recur within the same observation period (2 minutes). We do not analyse recurrent events in this tutorial. More information about recurrent events analysis can be found in REF and REF...

In secion A of the Supplementary Material we visualize the types of time-to-event data that can be obtained in these RT tasks.

## 2.2 Right censoring vs data trimming

## 2.3 Discrete vs continuous time units

All measurements of duration are discrete in nature. However, when the temporal resolution is high relative to the duration of the observation window, 
- interest in the shape of the hazard function <-> Cox
- non-smooth functions <-> parametric
- definition hazard prob and logistic regression

## 2.4 Discrete-time Hazard functions and conditional accuracy fun

def hazard

def survival

def cumulative distribution

def prob mass function

def conditional accuracz function

## 2.5 Number of time bins, repeated measures, and samples

## 2.6 Bayesian vs. frequentist approach

-------------------

We recommend several excellent textbooks for a comprehensive background context to EHA [@singerAppliedLongitudinalData2003; @allisonSurvivalAnalysisUsing2010] and for a more general introduction to understanding regression equations [@gelmanRegressionOtherStories2020; @winterStatisticsLinguistsIntroduction2019]. Our focus here is not on providing a detailed account of the underlying regression equations, since this topic has been comprehensively covered many times before. Instead, we want to provide an intuition regarding how EHA works in general, as well as in the context of experimental psychology. As such, we only supply regression equations in section D of the Supplemental Material.

## 2.1 Basic features of event history analysis

In EHA, the definition of hazard and the type of models employed depend on whether one is using continuous or discrete time units.

Since our focus here is on hazard models that use discrete time units, we describe that approach. After dividing time in discrete, contiguous time bins indexed by t (e.g., t = 1:10 time bins), let RT be a discrete random variable denoting the rank of the time bin in which a particular person's response occurs in a particular experimental condition. For example, the first response might occur at 546 ms and it would be in time bin 6 (any RTs from 501 ms to 600 ms).

Discrete-time EHA focuses on the discrete-time hazard function of event occurrence and the discrete-time survivor function (Figure 2). The equations that define both of these functions are reported in section A of the Supplemental Material.

The discrete-time hazard function gives you, for each time bin, the probability that the event occurs (sometime) in bin t, given that the event does not occur in previous bins. In other words, it reflects the instantaneous risk that the event occurs in the current bin, given that it has not yet occurred in the past, i.e., in one of the prior bins. In contrast, the discrete-time survivor function cumulates the bin-by-bin risks of event *non*occurrence to obtain the survival probability, the probability that the event occurs after bin t. In other words, the survivor function gives you for each time bin the likelihood that the event occurs in the future, i.e., in one of the subsequent time bins.

(ref:descr-plot2-caption) Discrete-time hazard and survivor functions. Discrete time-to-event data were simulated for 200 trials of 1 experimental condition. Error bars represent +/- 1 standard error of the respective proportion. While the hazard function is the vehicle for inferring the time course of cognitive processes, the survival probability S(t-1) can help to qualify or provide context to the interpretation of the hazard probability h(t). For example, the high hazard of .60 = h(t=6) is experienced only by 44 percent of the trials, as S(t=5) = .44. 
Because the survivor function is a decreasing function of time, the error bars in later parts of the hazard function will always be wider and less precise compared to earlier parts.

```{r plot2, fig.cap = "(ref:descr-plot2-caption)", out.width="80%"}
knitr::include_graphics("../sims/figures/haz_surv_single.jpeg")
```

## 2.3 Event history analysis in the context of experimental psychology

To make EHA more relevant to researchers studying cognitive psychology and cognitive neuroscience, in this section we provide a relevant worked example and consider implications that are relevant to that domain of research.

### 2.3.1 A worked example

In the context of experimental psychology, it is common for participants to be presented with either a 1-button detection task or a discrimination task. For example, a task may involve choosing between two response options with only one of them being correct. For such two-choice RT data, the discrete-time EHA of the RT data (hazard and survivor functions) can be extended with a discrete-time SAT analysis of the timed accuracy data. Specifically, the hazard function of event occurrence can be extended with the discrete-time conditional accuracy function, which gives you the probability that a response is correct given that it is emitted in time bin t [@kantowitzInterpretationReactionTime2021; @wickelgrenSpeedaccuracyTradeoffInformation1977; @allisonSurvivalAnalysisUsing2010]. We refer to this extended (hazard + conditional accuracy) analysis for choice RT data as EHA/SAT.

Integrating results between hazard and conditional accuracy functions for choice RT data can be informative for understanding psychological processes. To illustrate, we consider a hypothetical choice RT example that is inspired by real data [@panisWhatShapingRT2016], but simplified to make the main point clearer (Figure 3). In a standard priming paradigm, there is a prime stimulus (e.g., an arrow pointing left or right) followed by a target stimulus (another arrow pointing left or right). The prime can then be congruent or incongruent with the target.

Figure 3 shows that the early upswing in hazard is equal for both priming conditions (Figure 3, upper panel), and that early emitted responses are always correct in the congruent condition and always incorrect in the incongruent condition (Figure 3, lower panel). These results show that for short waiting times (< bin 6), responses always follow the prime (and not the target, as instructed). During time bin 6 the target-triggered response channel is activated and causes response competition -- ca(6) = .5 -- and a lower hazard probability in the incongruent condition. For waiting times of 600 ms or more, the hazard of response occurrence is lower in incongruent compared to congruent trials, and all responses emitted in these late bins are correct.

This joint pattern of results is interesting because it can provide meaningfully different conclusions about psychological processes compared to conventional analyses, such as computing mean-average RT and accuracy across trials. Mean-average RT would only represent the overall ability of cognition to overcome interference, on average, across trials. For instance, if mean-average RT was higher in incongruent than congruent trials, one may conclude that cognitive mechanisms that support interference control are working as expected across trials, and are indexed by each recorded response. But such a conclusion is not supported when the effects are explored over a timeline. Instead, the psychological conclusion is much more nuanced and suggests that multiple states start, stop and possibly interact over a particular temporal window.

(ref:descr-plot3-caption) Discrete-time hazard and conditional accuracy functions. Discrete time-to-event and conditional accuracy data were simulated for 1000 trials for each of two priming conditions (congruent and incongruent prime stimuli). Error bars represent +/- 1 standard error of the respective proportion. Bin width equals 100 ms.

```{r plot3, fig.cap = "(ref:descr-plot3-caption)", out.width="80%"}
#knitr::include_graphics("../sims/figures/haz_acc_single.jpeg")
```

Unlocking the temporal states of cognitive processes can be revealing for theory development and the understanding of basic psychological processes. Possibly more importantly, however, is that it simultaneously opens the door to address many new and previously unanswered questions. Do all participants show similar temporal states or are there individual differences? Do such individual differences extend to those individuals that have been diagnosed with some form of psychopathology? How do temporal states relate to brain-based mechanisms that might be studied using other methods from cognitive neuroscience? And how much of theory in cognitive psychology would be in need of revision if mean-average comparisons were supplemented with a temporal states approach?

### 2.3.2 Implications for designing experiments

Performing EHA in experimental psychology has implications for how experiments are designed. Indeed, if trials are categorised as a function of when responses occur, then each time bin will only include a subset of the total number of trials. For example, let's consider an experiment where each participant performs 2 conditions and there are 100 trial repetitions per condition. Those 100 trials must be distributed in some manner across the chosen number of bins.

In such experimental designs, since the number of trials per condition are spread across bins, it is important to have a relatively large number of trial repetitions per participant and per condition. Accordingly, experimental designs using this approach typically focus on factorial, within-subject designs, in which a large number of observations are made on a relatively small number of participants (so-called small-*N* designs). This approach emphasizes the precision and reproducibility of data patterns at the individual participant level to increase the inferential validity of the design [@bakerPowerContoursOptimising2021; @smithSmallBeautifulDefense2018].

In contrast to the large-*N* design that typically average across many participants without being able to scrutinize individual data patterns,  small-*N*  designs  retain  crucial  information  about  the  data  patterns of individual observers. This can be advantageous whenever participants differ systematically in their strategies or in the time courses of their effects, so that averaging them would lead to misleading data patterns. Note that because statistical power derives both from the number of participants and from the number of repeated measures per participant and condition, small-*N* designs can still achieve what are generally considered acceptable levels of statistical power, if they have a sufficient amount of data overall [@bakerPowerContoursOptimising2021; @smithSmallBeautifulDefense2018].

# 3. An overview of the general analytical workflow

Although the focus is on EHA/SAT, we also want to briefly comment on broader aspects of our general analytical workflow, which relate more to data science and data analysis workflows.

## 3.1 Data science workflow and descriptive statistics

We perform data wrangling following tidyverse principles and a functional programming approach [@wickhamDataScienceImport2023].
In short, functional programming means that you avoid writing your own loops and instead use functions that have been built and tested by others.
In addition, we also supply a set of custom-built functions, which make the process of data wrangling in the context of data preparation and descriptive statistics a lot quicker and more efficient.

## 3.2 Inferential statistical approach

Our lab adopts an estimation approach to multilevel regression [@kruschkeBayesianNewStatistics2018; @winterStatisticsLinguistsIntroduction2019], which is heavily influenced by the Bayesian framework as suggested by Richard McElreath [@mcelreathStatisticalRethinkingBayesian2020; @kurzStatisticalRethinkingSecondEd2023]. We also use a "keep it maximal" approach to specifying varying (or random) effects [@barrRandomEffectsStructure2013]. This means that wherever possible we include varying intercepts and slopes per participant.
To make inferences, we use two main approaches. We compare models of different complexity, using information criteria (e.g., WAIC) and cross-validation (e.g., LOO), to evaluate out-of-sample predictive accuracy [@mcelreathStatisticalRethinkingBayesian2020]. We also take the most complex model and evaluate key parameters of interest using point and interval estimates.

## 3.3 Implementation

We used `r my_r_citation$r` for all reported analyses. The content of the tutorials, in terms of EHA and multilevel regression modelling, is mainly based on @allisonSurvivalAnalysisUsing2010, @singerAppliedLongitudinalData2003, @mcelreathStatisticalRethinkingBayesian2020, @heiss2021,  @kurzAppliedLongitudinalDataAnalysis2023, and @kurzStatisticalRethinkingSecondEd2023.

`r my_r_citation$pkgs` <!-- create footnote -->

@@

@@

The next step is to set up the life table using our custom function setup_lt(), calculate the conditional accuracies using our custom function calc_ca(), add the ca(t) estimates to the life table using our custom function join_lt_ca(), and then plot the descriptive statistics using our custom function plot_eha(). When creating the plots, some warning messages will likely be generated, like these:

* Removed 2 rows containing missing values or values outside the scale range (`geom_line()`). 
* Removed 2 rows containing missing values or values outside the scale range (`geom_point()`). 
* Removed 2 rows containing missing values or values outside the scale range (`geom_segment()`).

The warning messages are generated because some bins have no hazard and ca(t) estimates, and no error bars. They can thus safely be ignored.
One can now inspect different aspects, including the life table for a particular condition of a particular subject, and a plot of the different functions for a particular participant. 
In general, it is important to visually inspect the functions first for each participant, in order to identify individuals that may be guessing (e.g., a flat conditional accuracy function at .5 indicates that someone is just guessing), outlying individuals, and/or different groups with qualitatively different behavior.

@@

@@

## 4.5 Tutorial 3a: Fitting Frequentist hazard models

In this fifth tutorial we illustrate how to fit a multilevel regression model to RT data in the frequentist framework, for the data used in Tutorial 1a. The general process is similar to that in Tutorial 2a, except that there are no priors to set.

Again, to keep the tutorial concise, we only fit model M1i (see Tutorial 2a) using the function glmer() from the R package lme4. Alternatively, one could also use the function glmmPQL() from the R package MASS [@ripleyMASSSupportFunctions2024]. The resulting hazard model is called M1i_f with the appended "_f" denoting a frequentist model.

In Figure 11 we compare the parameter estimates from the Bayesian regression model M1i with those from the frequentist model M1i_f.

>>>>

# R references
# r_refs(file = "r-references.bib", append=F) # append=F prevents multiple copies of the same citation
# my_r_citation <- cite_r(file = "r-references.bib", footnote = TRUE)
```

@@

As a simple illustration, Figure 1 summarises simulated data that shows how comparing means between two conditions can conceal the shapes of the underlying RT and accuracy distributions. Indeed, compared to the aggregation of data across trials (Figure 1A), a distributional approach offers the possibility to reveal the time course of psychological states (Figure 1B). For example, Figure 1B shows a first state (up to 400 ms after target onset) for which the early upswing in hazard is equal for both conditions, and the emitted responses are always correct in condition 1 and always incorrect in condition 2. In a second state (400 to 500 ms), hazard is higher in condition 1, and conditional accuracies are close to .5 in both conditions. In a third state (>500 ms), the effect disappears in hazard, and all conditional accuracies are equal to 1.

Why does this matter for research in psychology? For many psychological questions, the estimation of such "temporal states" information can be theoretically meaningful by leading to more fine-grained understanding of psychological processes and by adding a relatively under-used dimension -- the passage of time -- to the theory building toolkit. Thus, a distributional approach permits different kinds of questions to be asked, different inferences to be made, and it holds the potential to better discriminate between different theoretical accounts of psychological and/or brain-based processes.

(ref:descr-fig1-caption) Simulated data showing mean performance versus distributional analyses. (A) The mean RT (top) and overall accuracy (bottom) for two conditions are plotted. (B) The discrete-time hazard functions (top) and conditional accuracy functions (bottom) are plotted for the same data. The first second after target stimulus onset (time zero) is divided in ten bins of 100 ms. The first bin is (0,100], the last bin is (900,1000]. Note that the hazard and conditional accuracy estimates are plotted at the endpoint of each time bin. The definitions of discrete-time hazard and conditional accuracy are further explained in section 2. Error bars represent +/- 1 standard error of the mean (A) or proportion (B).

@@

## 1.2 Aims

Our ultimate aim in this paper is twofold: first, we want to convince readers of the many benefits of using EHA when dealing with psychological RT data, and second, we want to provide a set of practical tutorials, which provide step-by-step instructions on how you actually perform a discrete-time EHA on RT data, as well as a complementary discrete-time speed-accuracy tradeoff (SAT) analysis on timed accuracy data in case of choice RT data.

Even though EHA is a widely used statistical tool and there already exist many excellent reviews [e.g., @singerAppliedLongitudinalData2003; @allisonDiscreteTimeMethodsAnalysis1982; @blossfeldTechniquesEventHistory2002; @hosmerAppliedSurvivalAnalysis2011; @teachmanAnalyzingSocialProcesses1983; @box-steffensmeierEventHistoryModeling2004] and tutorials [e.g., @allisonSurvivalAnalysisUsing2010; @landesIntroductionEventHistory2020], we are not aware of any tutorials that are aimed specifically at psychological RT (+ accuracy) data, and which provide worked examples of the key data processing and Bayesian multilevel regression modelling steps. Set within this context, our overall aim is to introduce a set of tutorials, which explain **how** to do such analyses in the context of experimental psychology, rather than repeat in any detail **why** you may do them. Therefore, we hope that our tutorials will provide a pathway for research avenues in experimental psychology that have the potential to benefit from using EHA in the future.

## 1.3 Structure

In what follows, the paper is organised in three main sections. In Section 2, we provide a brief overview of EHA to orient the reader to the basic concepts that we will use throughout the paper and why such an approach might be relevant for research in experimental psychology. In Section 3, we outline a series of tutorials, which are written in the R programming language and publicly available on our Github page (https://github.com/sven-panis/Tutorial_Event_History_Analysis), along with all of the other code and material associated with the project. The tutorials provide hands-on, concrete examples of key parts of the analytical process, such as data wrangling, model fitting and planning future studies, so that others can apply EHA to their own time-to-event data measured in RT tasks. In Section 4, we discuss the strengths and weaknesses of the approach for researchers in experimental psychology.

# 2. What is event history analaysis and why is it relevant to research in experimental psychology?

## 2.1 A brief introduction to event history analysis

EHA is a statistical approach to study the occurrence and timing of events, such as disease onset, marriages, arrests, and job terminations (Allison, 2010). In this section, we want to provide an intuition regarding how EHA works in general, as well as in the context of experimental psychology. For those who want more detailed treatment of EHA and/or regression equations, we refer the reader to several excellent textbooks on these topics [@singerAppliedLongitudinalData2003; @allisonSurvivalAnalysisUsing2010; @gelmanRegressionOtherStories2020; @winterStatisticsLinguistsIntroduction2019]. We also supply all relevant regression equations in section D of the Supplemental Material.

### 2.1.1 Terminolgy and minimum requirements for EHA

To avoid possible confusion in terminology used, it is worth noting that EHA is frequently known by various labels, such as survival analysis, hazard analysis, duration analysis, failure-time analysis, and transition analysis (Singer & Willett, 2003). In this paper, we choose to use the term EHA throughout.

In terms of minimum requirements to apply EHA, one must be able to:

1. define an event of interest that represents a qualitative change - a transition from one discrete state to another - that can be situated in time (e.g., a button press, a saccade onset, a fixation offset, etc.);

@@

### 2.1.2 Types of EHA

There are different types of EHA.
For example, the definition of hazard and the type of models employed depend on whether one is using continuous or discrete time units. 
As a lab, and mainly for practical reasons, we have much more experience using discrete time EHA, and that is the approach that we describe and focus on in this paper. 
This choice may seem counter-intuitive, given that RT is typically treated as a continuous variable. 
However, continuous forms of EHA require much more data to estimate the continuous-time hazard (rate) function well. 
Thus, by trading a bit of temporal resolution for a lower number of trials, discrete-time methods seem ideal for dealing with typical psychological time-to-event data sets for which there are less than ~200 trials per condition per experiment.
Moreover, as indicated by @allisonSurvivalAnalysisUsing2010, learning discrete-time EHA methods first will help in learning continuous-time methods, so it seems like a good starting point.

To apply discrete time EHA, one divides time in discrete, contiguous time bins indexed by t (e.g., t = 1:10 time bins; Figure 1B).
Then let t be a discrete random variable denoting the rank of the time bin in which a particular person’s response occurs in a particular experimental condition. 
For example, a response on a single trial might occur at 546 ms and it would be in time bin 6 (any RTs from 501 ms to 600 ms).
One then calculates the discrete-time hazard function of event occurrence.
The discrete-time hazard function gives you, for each time bin, the probability that the event occurs (sometime) in
bin t, given that the event does not occur in previous bins. 
In other words, it reflects the instantaneous likelihood that the event occurs in the current bin, given that it has not yet
occurred in the past, i.e., in one of the prior bins.

In the context of experimental psychology, it is often (but not always), the case that responses can be classified as correct or incorrect.
In those cases, one can also calculate the conditional accuracy function (Figure B).
The conditional accuracy function gives you for each time bin the probability that a response is correct given that it is emitted in time bin t [@kantowitzInterpretationReactionTime2021; @wickelgrenSpeedaccuracyTradeoffInformation1977; @allisonSurvivalAnalysisUsing2010]. 
The ca(t) function is also known as the micro-level speed-accuracy tradeoff (SAT) function.
We refer to this extended (hazard + conditional accuracy) analysis for choice RT data as EHA/SAT.

## 2.2 Benefits of event history analysis for research in experimental psychology

Statisticians and mathematical psychologists recommend focusing on the hazard function when analyzing time-to-event data for various reasons (REF?? - Sven please cite a relevant REF here). 
We do not cover these benefits in detail here, as these are more general topics that have been covered elsewhere in textbooks (section F of the Supplemental Material). 
Instead, here we focus on the benefits as we see them for common research programmes in experimental psychology.

We highlight three benefits that we think are relevant to the domain of experimental psychology. 
First, as illustrated in Figure 1, compared to averaging data across trials, integrating results between hazard and conditional accuracy functions for choice RT data can be informative for understanding psychological processes, in terms of inferences about cognition and theoretical development.
As such, the approach permits different kinds of questions to be asked, different inferences to be made, and it holds the potential to discriminate between theoretical accounts of psychological and/or brain-based processes. 
For example, what kind of theory or theories could account for the effects reported in Figure 1B? 
Are there new auxiliary assumptions that theories need to adopt? 
And are there new experiments that need to be performed to test the novel predictions that follow from these analyses?

Second, compared to more conventional analytical approaches, EHA uses more of the data and deals with msising data differently.
It is conventional with RT data to either (a) use a response deadline and discard all trials without a response, or (b) wait in each trial until a response occurs and then apply data trimming techniques, i.e., discarding too short or too long RTs before calculating a mean RT (REF). 
Discarding data can introduce biases, however.
Rather than treat non-responses as missing data, EHA treats such trials as *right-censored* observations on the variable RT, because all we know is that RT is greater than some value. 
Right-censoring is a type of missing data problem and a nearly universal feature of survival data including RT data. 
For example, if the censor time was 1 second, then some trials result in observed event times (those with a RT below 1 second), while the other trials result in response times that are right-censored at 1 second.
The use of right-censoring in EHA, therefore, presents a analytical strength of the approach compared to many common approaches in experimental psychology.

Third, the approach is generalisable and applicable to many tasks that are commonly used in experimental psychology, such as detection, discrimination and bistable percpetion tasks (Supp Fig X). 
EHA can also be used with a range of common experimental manipualtions, such as stimulus-onset-asynchrony (Supp Fig X). 
The upshot is that one general analytical approach, which holds several potential advantages, is widely applicable to many substantive use-cases in the domain of experimental psychology.

## 2.3 Implications for research design in experimental psychology

Performing EHA in experimental psychology has implications for how experiments are designed. 
More specifically, we consider two implications that researchers will need to consider when using discrete time EHA.

First, since the number of trials per condition are spread across bins, it is important to have a relatively large number of trial repetitions per participant and per condition. Accordingly, experimental designs using this approach typically focus on factorial, within-subject designs, in which a large number of observations are made on a relatively small number of participants (so-called small-*N* designs). This approach emphasizes the precision and reproducibility of data patterns at the individual participant level to increase the inferential validity of the design [@bakerPowerContoursOptimising2021; @smithSmallBeautifulDefense2018]. Note that because statistical power derives both from the number of participants and from the number of repeated measures per participant and condition, small-*N* designs can still achieve what are generally considered acceptable levels of statistical power, if they have a sufficient amount of data overall [@bakerPowerContoursOptimising2021; @smithSmallBeautifulDefense2018].

Second, the width of each time bin will need to be determined. 
For instance, in Figure 1B we chose 100ms in an arbitrary manner.
In reality, however, bin width will need to be set by considering a number of factors simultabeously.
For example, [[Sven - add a few sentences here]]

@@

We used r my_r_citation$r for all reported analyses. The content of the tutorials, in terms of EHA and multilevel regression modelling, is mainly based on @allisonSurvivalAnalysisUsing2010, @singerAppliedLongitudinalData2003, @mcelreathStatisticalRethinkingBayesian2020, @heiss2021,  @kurzAppliedLongitudinalDataAnalysis2023, and @kurzStatisticalRethinkingSecondEd2023.

@@

The next step is to set up the life table using our custom function setup_lt(), calculate the conditional accuracies using our custom function calc_ca(), add the ca(t) estimates to the life table using our custom function join_lt_ca(), and then plot the descriptive statistics using our custom function plot_eha(). 
One can now inspect different aspects, including the life table for a particular condition of a particular subject, and a plot of the different functions for a particular participant. 
In general, it is important to visually inspect the functions first for each participant, in order to identify individuals that may be guessing (e.g., a flat conditional accuracy function at .5 indicates that someone is just guessing), outlying individuals, and/or different groups with qualitatively different behavior.

@@

In general, when fitting regression models, our lab adopts an estimation approach to multilevel regression [@kruschkeBayesianNewStatistics2018; @winterStatisticsLinguistsIntroduction2019], which is heavily influenced by the Bayesian framework as suggested by Richard McElreath [@mcelreathStatisticalRethinkingBayesian2020; @kurzStatisticalRethinkingSecondEd2023]. We also use a "keep it maximal" approach to specifying varying (or random) effects [@barrRandomEffectsStructure2013]. This means that wherever possible we include varying intercepts and slopes per participant.
To make inferences, we use two main approaches. We compare models of different complexity, using information criteria (e.g., WAIC) and cross-validation (e.g., LOO), to evaluate out-of-sample predictive accuracy [@mcelreathStatisticalRethinkingBayesian2020]. We also take the most complex model and evaluate key parameters of interest using point and interval estimates.

@@

<!-- ## 4.5 Tutorial 3a: Fitting Frequentist hazard models -->

<!-- In this fifth tutorial we illustrate how to fit a multilevel regression model to RT data in the frequentist framework, for the data used in Tutorial 1a. The general process is similar to that in Tutorial 2a, except that there are no priors to set.  -->

<!-- Again, to keep the tutorial concise, we only fit model M1i (see Tutorial 2a) using the function glmer() from the R package lme4. Alternatively, one could also use the function glmmPQL() from the R package MASS [@ripleyMASSSupportFunctions2024]. The resulting hazard model is called M1i_f with the appended "_f" denoting a frequentist model. -->

<!-- In Figure 11 we compare the parameter estimates from the Bayesian regression model M1i with those from the frequentist model M1i_f. -->
